A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-05-21 manuscript.qmd
revise more

For simpler statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups), user-friendly software for a priori sample size planning is readily available [@faulStatisticalPowerAnalyses2009; @champelyPackagePwr2018; @kelleyObtainingPowerObtaining2003; @kelleySampleSizePlanning2006]. 
However, these software packages are often not flexible enough to perform sample size planning for complex designs.

Existing approaches for sample size planning with GLMMs have almost exlusively focused on planning for power. 
Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022].
Formula-based methods rely on often complicated formulas that can be used to directly calculate power while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., how often the hypothesis test is significant for the simulated data).
Currently available formula-based software packages for power analysis with multilevel models often do not include GLMMs or are limited to very simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. 
A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. 
However, most of these tutorials focus on linear mixed models (LMMs) and the most common designs [but see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers faced with more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. 
The necessary presumptions for simulation-based power analysis with GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the correlation structure. The distributional assumption specifies the distributional family for the outcome variable (when conditioned on all fixed and random effects). Assumptions about the random effects include the assumption of normality (i.e., that the random effects follow a normal distribution) and the covariance structure among the random effects (i.e.,  if and how they are correlated). Interpreting these presumptions entails understanding the underlying presumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge [see the discussion in @maxwellSampleSizePlanning2008 and @kumleEstimatingPowerGeneralized2021].

@@

In the first six lines of the function definition, we set some default parameter values (which we will explain in the next section) and load the packages we use to manipulate and simulate data. 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus, the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows.
We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts.
The next two lines simulate that in $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`).
As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user.
With all design variables done, we are ready to simulate our model equation outlined in the last section.
The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects, and model parameters as assumed by our model.
We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`).
In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject makes an accurate diagnostic decision for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

>>>>

<!-- For less complex statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups), user-friendly software for a priori sample size planning is readily available [@faulStatisticalPowerAnalyses2009; @champelyPackagePwr2018; @kelleyObtainingPowerObtaining2003; @kelleySampleSizePlanning2006].  -->
<!-- However, these software packages are often not flexible enough to perform sample size planning for complex designs. -->

Existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on formulas to calculate power directly while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., how often the hypothesis test is significant for the simulated data). Available formula-based software packages for multilevel models often do not include GLMMs or are limited to very simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. However, most of these tutorials focus on linear mixed models (LMMs) and the common study designs [but see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers faced with more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. The prerequisites for simulation-based power analysis with GLMMs include making assumptions about the distributional form of the outcome variable, the random effects, and the correlation structure. The distributional assumption specifies the distributional family for the outcome variable. Assumptions about the random effects include the assumption of normality (i.e., that the random effects follow a normal distribution) and the covariance structure among the random effects (i.e., if and how they are correlated). Making these decisions requires understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in [@maxwellSampleSizePlanning2008] and [@kumleEstimatingPowerGeneralized2021]).

@@

In the first four lines of the function definition, we set some default parameter values (which we will explain in the next section) and load the packages we use to manipulate and simulate data. 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus, the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows.
We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts.
The next two lines simulate that in $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`).
As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user.
With all design variables done, we are ready to simulate our model equation outlined in the last section.
The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects, and model parameters as assumed by our model.
We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`).
In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject makes an accurate diagnostic decision for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.
